### 2014-08-03 Pero_Transcriptome.tex
1730

Deserts are widely considered one of Earth's harshest environments. Animals living in desert environments are forced to endure intense heat and drought, and in turn, species having evolved in these environments are likely ton have evolved specialised mechanisms that may enhance fitness. While living in deserts likely involves a large number of adaptive phenotypes, the ability to osmoregulate -- to maintain the proper water and electrolyte balance -- appears to be paramount \cite{Walsberg:2000uu}. Indeed, the maintenance of water balance in animals is one of the most important physiologic processes, and is critical to desert survival. Mammals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur. Understanding these mechanismsm will significantly enhance our understanding of the physiologic processes underlying osmoregulation in extreme environments, having implications for studies of human health, conservation, and climate change. \\

@@

To begin to understand how genes may underlie desert adaptation, I collected ~15 individuals from a single population \textit{P. eremicus} over a two year time period (2012-2013). These individuals were captured in live traps, then euthanized using isoflurane overdose and decapitation. Immediately post-mortem, the abdominal and pelvic organs were removed, cut in half (in the case of kidney), placed in RNAlater and flash frozen in liquid Nitrogen. Removal of the brain, with similar preservation techniques, followed that. Time from euthanasia to removal of all organs never exceeded five minutes. Samples were transferred to a -80C freezer at a later date. These procedures were approved by the University of California Berkeley Animal Care and Use Committee and follow guidelines established by the American Society of Mammalogy for the use of wild animals in research \cite{Sikes:2011dz}.

@@

From the filtered assembly, I extracted putative coding sequences using Transdecoder version 4ul2014 (\url{http://transdecoder.sourceforge.net/}). These putative protein coding sequences were annotated using default settings of the blastx algorithm \cite{Camacho:2009fc} against the SwissProt database downloaded on 1 March 2014. Because transcriptome assemblies typically contain non-coding elements (e.g. ncRNA) in addition to protein coding sequence, we annotated the entire filtered dataset using the NCBI nt dataset, downloaded on 1 March 2014. Lastly, because rapidly evolving nucleotide sequences may evade detection by blast algorithms, we used HMMER3 \cite{Wheeler:2013gj} to search for conserved protein domains contained in the Pfam database \cite{Punta:2012ko}. \\

To identify sequences unique to each tissue type, I mapped sequence reads from each tissue type to the reference assembly using bwa-mem. We estimated expression individually for the four tissues. Interesting patterns of expression, including instances where expression was limited to a single tissue type were identified. \\

@@

Transcriptome assembly for each tissue type was accomplished using the program Trinity. The raw assembly for brain, liver, testes and kidney contained 185425, 222096, 180233, and 514091 assembled sequences respectively. This assembly was filtered using a blast procedure with resulted in a final dataset containing 68331 brain-specific transcripts, 71041 liver-specific transcripts, 67340 testes-specific transcripts, and 113050 kidney-specific transcripts. Mapping the error-corrected adapter/quality trimmed reads to these datasets resulted in mapping 94.98\% (87.01\% properly paired) of brain-derived reads to the brain transcriptome, 96.07\% (88.13\% properly paired) of liver-derived reads to the liver transcriptome, 96.81\% (85.10\% properly paired) of testes-derived reads to the testes transcriptome, and 91.87\% (83.77\% properly paired) of kidney-derived reads to the kidney transcriptome. Together, these statistics suggest that the tissue-specific transcriptomes are of extremely high quality.  \\

%\vspace{5mm}
%\textbf{\hypertarget{Figure 1}{Figure 1}} \\
\centerline{\includegraphics[width=20.0\baselineskip]{/Users/macmanes/Documents/pero_transcriptome/venn.jpeg}}
\begin{quote}
\small{Figure 1. The Venn Diagram.}
\end{quote}

I then estimated gene expression on each of these tissue-specific datasets, which allowed me to understand expression patterns in the multiple tissues {\hypertarget{Figure 1}{Figure 1}}. After expression estimation, the filtered assemblies were concatenated together, and after removal of redundancy with cd-hit-est, 123,123 putative transcripts remained. From this filtered concatenated dataset, I extracted 71626 putative coding sequences (72Mb). Of these 71626 sequences, 38221 were complete exons (containing both start and stop codons), while other were either truncated at the 5-prime end (20239 sequences), 3-prime end (6445 sequences), or were internal (6721 sequencing having neither stop nor start codon).

>>>>

Deserts are widely considered one of Earth's harshest environments. Animals living in desert environments are forced to endure intense heat and drought, and in turn, species having evolved in these environments are likely ton have evolved specialised mechanisms that may enhance fitness. While living in deserts likely involves a large number of adaptive phenotypes, the ability to osmoregulate -- to maintain the proper water and electrolyte balance -- appears to be paramount \cite{Walsberg:2000uu}. Indeed, the maintenance of water balance in animals is one of the most important physiologic processes for all animals, desert inhabitants or not. Most animals are exquisitely sensitive to changes in osmolality, with slight derangement eliciting physiologic compromise.  When the loss of water exceeds dietary intake, dehydration - and in extreme cases, death - can occur, which suggests that there is strong selection for mechanisms supporting osmoregulation. Understanding these mechanisms will significantly enhance our understanding of the physiologic processes underlying osmoregulation in extreme environments, having implications for studies of human health, conservation, and climate change. \\

@@

To begin to understand how genes may underlie desert adaptation, I collected 16 individuals from a single population \textit{P. eremicus} over a two year time period (2012-2013). These individuals were captured in live traps, then euthanized using isoflurane overdose and decapitation. Immediately post-mortem, the abdominal and pelvic organs were removed, cut in half (in the case of kidney), placed in RNAlater and flash frozen in liquid Nitrogen. Removal of the brain, with similar preservation techniques, followed that. Time from euthanasia to removal of all organs never exceeded five minutes. Samples were transferred to a -80C freezer at a later date. These procedures were approved by the University of California Berkeley Animal Care and Use Committee and follow guidelines established by the American Society of Mammalogy for the use of wild animals in research \cite{Sikes:2011dz}.

@@

The filtered assemblies were annotated using default settings of the blastN algorithm \cite{Camacho:2009fc} against the Ensembl cDNA and ncRNA datasets described above, downloaded on 1 August 2014. Amongst other things, the Ensemble transcript identifiers were used in the analysis of gene ontology, conducted in the PANTHER package \cite{Mi:2004iv}. Next, because rapidly evolving nucleotide sequences may evade detection by blast algorithms, we used HMMER3 \cite{Wheeler:2013gj} to search for conserved protein domains contained in the dataset using the Pfam database \cite{Punta:2012ko}. Lastly, I extracted putative coding sequences using Transdecoder version 4Jul2014 (\url{http://transdecoder.sourceforge.net/})\\

To identify patterns of gene expression unique to each tissue type, I mapped sequence reads from each tissue type to the reference assembly using bwa-mem. We estimated expression individually for the four tissues using default settings of the software eXpress \cite{Roberts:2012dh}. Interesting patterns of expression, including instances where expression was limited to a single tissue type were identified and visualized. \\

@@

Transcriptome assembly for each tissue type was accomplished using the program Trinity. The raw assembly for brain, liver, testes and kidney contained 185425, 222096, 180233, and 514091 assembled sequences respectively. This assembly was filtered using a blastN procedure against the \textit{Mus} cDNA and ncRNA which resulted in a final dataset containing 68331 brain-specific transcripts, 71041 liver-specific transcripts, 67340 testes-specific transcripts, and 113050 kidney-specific transcripts. Mapping the error-corrected adapter/quality trimmed reads to these datasets resulted in mapping 94.98\% (87.01\% properly paired) of brain-derived reads to the brain transcriptome, 96.07\% (88.13\% properly paired) of liver-derived reads to the liver transcriptome, 96.81\% (85.10\% properly paired) of testes-derived reads to the testes transcriptome, and 91.87\% (83.77\% properly paired) of kidney-derived reads to the kidney transcriptome. Together, these statistics suggest that the tissue-specific transcriptomes are of extremely high quality.  \\

%\vspace{5mm}
%\textbf{\hypertarget{Figure 1}{Figure 1}} \\
\centerline{\includegraphics[width=15.0\baselineskip]{/Users/macmanes/Documents/pero_transcriptome/venn.jpeg}}
\begin{quote}
\small{Figure 1. The Venn Diagram.}
\end{quote}

I then estimated gene expression on each of these tissue-specific datasets, which allowed me to understand expression patterns in the multiple tissues ({\hypertarget{Figure 1}{Figure 1}}). After expression estimation, the filtered assemblies were concatenated together, and after removal of redundancy with cd-hit-est, 123,123 putative transcripts remained. From this filtered concatenated dataset, I extracted 71626 putative coding sequences (72Mb). Of these 71626 sequences, 38221 were complete exons (containing both start and stop codons), while other were either truncated at the 5-prime end (20239 sequences), 3-prime end (6445 sequences), or were internal (6721 sequencing having neither stop nor start codon).
